GMSE: an R package for generalised management strategy evaluation

### 2018-04-18 notebook/ms/SI3.Rmd
More edits to SI text

---
title: 'GMSE: an R package for generalised management strategy evaluation'
subtitle: Supporting Information 3
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: "[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK
[2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
linestretch: 1
bibliography: gmse.bib
biblio-style: apalike
link-citations: true
linkcolor: blue
---

```{r, echo = FALSE}
library(GMSE);
```

@@

```{r, echo = TRUE}
sim_1 <- gmse(manager_budget = 10000, user_budget = 10000, plotting = FALSE, 
              res_death_K = 93870, manage_target = 70000, RESOURCE_ini = 35000, 
              stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
              scaring = FALSE, lambda = 0.150, remove_pr = 0.122, time_max = 20, 
              res_death_type = 3, res_consume = 0.01, res_birth_K = 200000, 
              observe_type = 3, agent_view = 1);
```

@@

```{r, eval = TRUE, fig.height = 8, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit. This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 20 simulated time steps. Goose locations at the end of the simulation are shown in the upper left panel, while the upper right panel shows the same landscape broken down among the 80 farmers (upper 60% of the landscape in multiple colours), along with non-agricultural land (lower 40% of the landscape in blue). Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line). Horizontal red and blue dotted lines show landscape-level goose carrying capacity enacted on adult mortality and the manager's target for goose abundance, respectively. The organge line shows the total percent of landscape cells (including non-farmed cells) yield, as decreased by geese. The middle right panel shows this yield for each individual farmer, and for the public land (lower line in blue). The lower left panel shows the cost of culling for farmers, as set by the manager, and the lower right panel shows the total number of culls attempted by farmers over time."}
plot_gmse_results(res = sim_1$resource, obs = sim_1$observation, 
                  land = sim_1$land, agents = sim_1$agents, paras = sim_1$paras, 
                  ACTION = sim_1$action, COST = sim_1$cost);
```

@@

```{r, eval = TRUE, fig.height = 7, fig.width = 6, fig.cap = "Permissiveness that each manager exhibits for each farmer action (black lines) and the effort that each individual farmer puts into each action over time (coloured lines). Each panel row reports a different action (in decreasing order: scaring, culling, castration, feeding, and helping). The left axis shows the permissiveness that a manager has for the focal action (black line), which is calculated as 100 minus the percent of the manager's budget that is put into increasing the cost of the focal action. For example, if the manager puts all of their effort (total budget) into increasing the cost of culling, then permissiveness of culling is 0; if the manager puts no effort into increasing culling cost, then permissiveness of culling is 100. The right axis shows effort that farmers put into an action (coloured lines), which defined as the percentage of a farmer's budget put into a particular action (note, values might not add up to 100 because farmers are not forced to use their entire budget)."}
plot_gmse_effort(sim_1$agents, sim_1$paras, ACTION = sim_1$action, 
                 COST = sim_1$cost);
```

@@

```{r, echo = TRUE, fig.height = 8}
sim_2 <- gmse(manager_budget = 10000, user_budget = 10000, plotting = FALSE, 
              res_death_K = 93870, manage_target = 70000, RESOURCE_ini = 35000, 
              stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
              scaring = TRUE, lambda = 0.150, remove_pr = 0.122, time_max = 20, 
              res_death_type = 3, res_consume = 0.01, res_birth_K = 200000, 
              observe_type = 3, agent_view = 1);
```

@@

References
--------------------------------------------------------------------------------

>>>>

---
title: 'GMSE: an R package for generalised management strategy evaluation'
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;,
  Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes: null
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
subtitle: Supporting Information 3
biblio-style: apalike
---

```{r, echo = FALSE}
library(GMSE);
plot_gmse_effort <- function(agents, paras, ACTION, COST){

    cols      <- c("green", "indianred1", "indianred3", "deepskyblue1",
                   "deepskyblue2");

    users    <- dim(agents[[1]])[1];
    max_time <- length(ACTION);

    para_vec <- paras[1,];

    allowed  <- sum(para_vec[89:93]);

    scar_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
    cull_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
    cast_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
    feed_cst  <- matrix(data = 0, nrow = max_time, ncol = users);
    help_cst  <- matrix(data = 0, nrow = max_time, ncol = users);

    scar_act  <- matrix(data = 0, nrow = max_time, ncol = users);
    cull_act  <- matrix(data = 0, nrow = max_time, ncol = users);
    cast_act  <- matrix(data = 0, nrow = max_time, ncol = users);
    feed_act  <- matrix(data = 0, nrow = max_time, ncol = users);
    help_act  <- matrix(data = 0, nrow = max_time, ncol = users);

    scar_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
    cull_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
    cast_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
    feed_eff  <- matrix(data = 0, nrow = max_time, ncol = users);
    help_eff  <- matrix(data = 0, nrow = max_time, ncol = users);

    act_costs  <- matrix(data = 0, nrow = max_time, ncol = 5);
    pol_effort <- matrix(data = 0, nrow = max_time, ncol = 5);
    min_cost   <- para_vec[97];

    for(time in 1:max_time){
        #-- Scaring cost and actions
        scar_cst[time, 1] <- COST[[time]][3, 8, 1];
        for(user in 2:users){
            scar_cst[time, user] <- COST[[time]][1, 8, user];
        }
        scar_act[time, 1] <- ACTION[[time]][3, 8, 1] - min_cost;
        for(user in 2:users){
            scar_act[time, user] <- ACTION[[time]][1, 8, user];
        }
        scar_eff[time,] <- scar_act[time,]*scar_cst[time,]/agents[[time]][,17];
        #-- Culling cost and actions
        cull_cst[time, 1] <- COST[[time]][3, 9, 1];
        for(user in 2:users){
            cull_cst[time, user] <- COST[[time]][1, 9, user];
        }
        cull_act[time, 1] <- ACTION[[time]][3, 9, 1] - min_cost;
        for(user in 2:users){
            cull_act[time, user] <- ACTION[[time]][1, 9, user];
        }
        cull_eff[time,] <- cull_act[time,]*cull_cst[time,]/agents[[time]][,17];
        #-- Castration cost and actions
        cast_cst[time, 1] <- COST[[time]][3, 10, 1];
        for(user in 2:users){
            cast_cst[time, user] <- COST[[time]][1, 10, user];
        }
        cast_act[time, 1] <- ACTION[[time]][3, 10, 1] - min_cost;
        for(user in 2:users){
            cast_act[time, user] <- ACTION[[time]][1, 10, user];
        }
        cast_eff[time,] <- cast_act[time,]*cast_cst[time,]/agents[[time]][,17];
        #-- Feeding cost and actions
        feed_cst[time, 1] <- COST[[time]][3, 11, 1];
        for(user in 2:users){
            feed_cst[time, user] <- COST[[time]][1, 11, user];
        }
        feed_act[time, 1] <- ACTION[[time]][3, 11, 1] - min_cost;
        for(user in 2:users){
            feed_act[time, user] <- ACTION[[time]][1, 11, user];
        }
        feed_eff[time,] <- feed_act[time,]*feed_cst[time,]/agents[[time]][,17];
        #-- Helping cost and actions
        help_cst[time, 1] <- COST[[time]][3, 12, 1];
        for(user in 2:users){
            help_cst[time, user] <- COST[[time]][1, 12, user];
        }
        help_act[time, 1] <- ACTION[[time]][3, 12, 1] - min_cost;
        for(user in 2:users){
            help_act[time, user] <- ACTION[[time]][1, 12, user];
        }
        help_eff[time,] <- help_act[time,]*help_cst[time,]/agents[[time]][,17];
    }

    # -- Turn these into percentages:
    scar_eff <- scar_eff * 100;
    cull_eff <- cull_eff * 100;
    cast_eff <- cast_eff * 100;
    feed_eff <- feed_eff * 100;
    help_eff <- help_eff * 100;

    par(mfrow = c(5, 1), mar = c(0, 0, 0, 0), oma = c(6, 6, 4, 6));
    y1 <- 100;
    y2 <- 130;
    lcex  <- 1.5
    if(max_time < 100){
        lcex <- 0.9;
    }
    #---- Scaring
    if(para_vec[89] == 1){
        plot(x = 1:max_time, y = 100 - scar_eff[,1], type = "l", lwd = 2, 
             cex.axis = 1.5, xaxt = "n", ylim = c(0, y2), yaxt = "n");
        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
        par(new = TRUE);
        plot(x = 1:max_time, y = scar_eff[,2], type = "n", lwd = 2, 
             ylim = c(0, y2), xaxt="n", yaxt="n", 
             cex.axis = 1.5);
        for(stakeholder in 2:users){
            points(x = 1:max_time, y = scar_eff[,stakeholder], type = "l", 
                   lwd = 1, col = cols[1]);
        }
        axis(side = 4, at = c(50, 100),  labels = c(50, 100), cex.axis = 1.5);
        legend(x = 1, y = 135, fill = cols[1:5], horiz = TRUE,
               legend = c("scaring", "culling", "castration", "feeding", 
                          "helping"), bty = "n", cex = lcex);
        abline(h = 105, lwd = 2);
    }else{
        plot(x = 1:max_time, y = scar_eff[,1], type = "n", xaxt= "n", 
             yaxt = "n", ylim = c(0, y2));
        text(x = 0, y = 88, cex = 2, labels = "No scaring allowed",
             pos = 4);
        abline(h = 105, lwd = 2);
        legend(x = 1, y = 135, fill = cols[1:5], horiz = TRUE,
               legend = c("scaring", "culling", "castration", "feeding", 
                          "helping"), bty = "n", cex = lcex);
    }
    #---- Culling
    if(para_vec[90] == 1){
        plot(x = 1:max_time, y = 100 - cull_eff[,1], type = "l", lwd = 2, 
             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
        par(new = TRUE);
        plot(x = 1:max_time, y = cull_eff[,2], type = "n", lwd = 2, 
             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
        for(stakeholder in 2:users){
            points(x = 1:max_time, y = cull_eff[,stakeholder], type = "l", 
                   lwd = 1, col = cols[2]);
        }
        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
    }else{
        plot(x = 1:max_time, y = cull_act[,1], type = "n", xaxt= "n", 
             yaxt = "n", ylim = c(0, 100));
        text(x = 0, y = 90, cex = 2, labels = "No culling allowed",
             pos = 4);
    }
    #---- Castrating
    if(para_vec[91] == 1){
        plot(x = 1:max_time, y = 100 - cast_eff[,1], type = "l", lwd = 2, 
             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
        par(new = TRUE);
        plot(x = 1:max_time, y = cast_eff[,2], type = "n", lwd = 2, 
             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
        for(stakeholder in 2:users){
            points(x = 1:max_time, y = cast_eff[,stakeholder], type = "l", 
                   lwd = 1, col = cols[3]);
        }
        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
    }else{
        plot(x = 1:max_time, y = cast_act[,1], type = "n", xaxt= "n", 
             yaxt = "n", ylim = c(0, 100));
        text(x = 0, y = 90, cex = 2, labels = "No castration allowed",
             pos = 4);
    }
    #---- Axes labels
    mtext("Manager's permissiveness of user action", 
          side = 2, line = 3.5, cex = 1.5, col = "black");
    mtext("Percentage of user action effort expended", 
          side = 4, line = 3.5, cex = 1.5, col = "black");
    #---- Feeding
    if(para_vec[92] == 1){
        plot(x = 1:max_time, y = 100 - feed_eff[,1], type = "l", lwd = 2, 
             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
        axis(side = 2, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
        par(new = TRUE);
        plot(x = 1:max_time, y = feed_eff[,2], type = "n", lwd = 2, 
             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
        for(stakeholder in 2:users){
            points(x = 1:max_time, y = feed_eff[,stakeholder], type = "l", 
                   lwd = 1, col = cols[4]);
        }
        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
    }else{
        plot(x = 1:max_time, y = feed_act[,1], type = "n", xaxt= "n", 
             yaxt = "n", ylim = c(0, 100));
        text(x = 0, y = 90, cex = 2, labels = "No feeding allowed",
             pos = 4);
    }
    #---- Helping
    if(para_vec[93] == 1){
        plot(x = 1:max_time, y = 100 - help_eff[,1], type = "l", lwd = 2, 
             cex.axis = 1.5, xaxt = "n", ylim = c(0, y1), yaxt = "n");
        axis(side = 2, at = c(50, y2), labels = c(50, 100), cex.axis = 1.5);
        par(new = TRUE);
        plot(x = 1:max_time, y = help_eff[,2], type = "n", lwd = 2, 
             ylim = c(0, y1), xaxt="n", yaxt="n", cex.axis = 1.5);
        for(stakeholder in 2:users){
            points(x = 1:max_time, y = help_eff[,stakeholder], type = "l", 
                   lwd = 1, col = cols[5]);
        }
        axis(side = 4, at = c(50, 100), labels = c(50, 100), cex.axis = 1.5);
    }else{
        plot(x = 1:max_time, y = help_act[,1], type = "n", cex.axis = 1.5,
             yaxt = "n", ylim = c(0, 100));
        text(x = 0, y = 90, cex = 2, labels = "No helping offspring allowed",
             pos = 4);
    }    
    mtext("Time step", side = 1, line = 3.5, cex = 1.5, col = "black");
}
```

@@

```{r, echo = TRUE, eval = FALSE}
sim_1 <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
             manage_target = 70000, RESOURCE_ini = 35000, plotting = FALSE, 
             stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
             scaring = FALSE, lambda = 0.275, remove_pr = 0.122, time_max = 40, 
             res_death_type = 3, res_consume = 0.02, res_birth_K = 200000, 
             observe_type = 3, agent_view = 1, converge_crit = 0.1);
```

```{r}
sim_1 <- gmse(time_max = 30, plotting = FALSE);
```

@@

```{r, eval = TRUE, fig.height = 7, fig.cap = "Results of a GMSE simulation using parameters estimated for Taiga Been Geese Central Management Unit. This example includes 80 farmers whose objective is to maximise their agricultural output, and one manager whose objective is to keep geese at a target abundance, over 20 simulated time steps. Goose locations at the end of the simulation are shown in the upper left panel, while the upper right panel shows the same landscape broken down among the 80 farmers (upper 60% of the landscape in multiple colours), along with non-agricultural land (lower 40% of the landscape in blue). Actual goose abundance is shown in the middle left panel (black solid line), along with its estimate by the manager (blue solid line). Horizontal red and blue dotted lines show landscape-level goose carrying capacity enacted on adult mortality and the manager's target for goose abundance, respectively. The organge line shows the total percent of landscape cells (including non-farmed cells) yield, as decreased by geese. The middle right panel shows this yield for each farmer, and for the public land (lower line in blue). The lower left panel shows the cost of culling for farmers, as set by the manager, and the lower right panel shows the total number of culls attempted by farmers over time."}
plot_gmse_results(res = sim_1$resource, obs = sim_1$observation, 
                  land = sim_1$land, agents = sim_1$agents, paras = sim_1$paras, 
                  ACTION = sim_1$action, COST = sim_1$cost);
```

@@

```{r, eval = TRUE, echo = TRUE, fig.height = 7, fig.width = 6, fig.cap = "Permissiveness that each manager exhibits for each farmer action (black lines) and the effort that each individual farmer puts into each action over time (coloured lines). Each panel row reports a different action (in decreasing order: scaring, culling, castration, feeding, and helping). The left axis shows the permissiveness that a manager has for the focal action (black line), which is calculated as 100 minus the percent of the manager's budget that is put into increasing the cost of the focal action. For example, if the manager puts all of their effort (total budget) into increasing the cost of culling, then permissiveness of culling is 0; if the manager puts no effort into increasing culling cost, then permissiveness of culling is 100. The right axis shows effort that farmers put into an action (coloured lines), which defined as the percentage of a farmer's budget put into a particular action (note, values might not add up to 100 because farmers are not forced to use their entire budget)."}
plot_gmse_effort(sim_1$agents, sim_1$paras, ACTION = sim_1$action, 
                 COST = sim_1$cost);
```

@@

```{r}
sim_2 <- gmse(time_max = 30, scaring = TRUE, plotting = FALSE);
```

```{r, echo = TRUE, fig.height = 6, eval = FALSE}
sim_2 <- gmse(manager_budget = 10000, user_budget = 10000, res_death_K = 93870, 
             manage_target = 70000, RESOURCE_ini = 35000, plotting = FALSE, 
             stakeholders = 80, land_ownership = TRUE, public_land = 0.4, 
             scaring = TRUE, lambda = 0.275, remove_pr = 0.122, time_max = 40, 
             res_death_type = 3, res_consume = 0.02, res_birth_K = 200000, 
             observe_type = 3, agent_view = 1, converge_crit = 0.1);
```

@@
